Sampling requirements to detect ecosystem change

### 2020-03-31 R/manuscript_draft.Rmd
added figure caption and corrected small spelling mistakes

Environmental monitoring is one of the core components to modern ecosystem research and management [@McDonald-Madden2010a, @White2019d, @Lindenmayer2020a]. Within an adaptive management framework, monitoring is needed for both learning about the system under study and assessing the effectiveness of management interventions [@Lovett2007]. Increasingly, long-term monitoring programs, like the Long Term Ecological Research (LTER) Network in the USA, are becoming available [@Maguran2010a]. However, environmental monitoring is still often expensive and time-consuming, especially when further procesing is needed following sample collection (citations). Thus, for many fields there is a disparity between the amount of data that can be aquired and stored, and the ultimate number of samples that can be processed. Therefore, monitoring programs need to be designed in such a way to address the question of interest while using limited resources efficiently [@Legg2006a, @McDonald-Madden2010a, @Lindenmayer2020a].<!-- processing methods that could allow to extract the most information from that data. \rosalie{I feel like it could be a sensitive subject (we don't want to attack our peers and the way they do science), but there is definitely some truths. With some restoration work going on at the lab, there were some emails exchange about throwing away samples that were collected in 2015 and never looked at. If we could find the correct way to talk about this issue (without being judgemental, I'm sure it'd be interesting to many!).}

@@

Accounting for key considerations in the design of a monitoring program is necessary to detect long-term ecological change. The specifics of the monitoring program will determine the power with which a question of interest can be addressed. For example, @White2019d found that 72\% of vertebrate populations required at least 10 years of monitoring to detect significant changes in the population size over time. The specific number of years required depended on the species biology and the detection method used [@White2019d]. Other work has focused on the frequency of monitoring [@Wauchope2019a], showing that if a significant trend is observed from a dataset with limited temporal perspective, it is likely to reliably describe qualitatively the complete trend (increase or decline), but is unlikely to provide an accurate quantitave change in population. Further research investigated the impact of allocating monitoring resources spatially versus temporally [@Rhodes2011a, @Weiser2019a], and the ability to use citizen science to increase monitoring [@Weiser2020a]. Lastly, both the ecological and economical costs of failing to detect a true trend (type II error) have to be weighed against the risks of false (type I error) detection [@Mapstone1995a]. The approach to optimal sampling is thus a topical subject of research, as under limited budgets, monitoring programs need to be designed to be cost-effective [@Caughlan2001a, @Grantham2008b, @Bennett2016a].

@@

We detect changepoints with the function _e.divisive_ in the R package _ecp_ [@James2019]. There are several methods available to detect changepoint (reviewed in @James2014); _e.divisive_ is a divisive hierarchical estimation algorithm for multiple change point analysis. We chose this method because it is able to perform multiple change point analysis for both uni- and multi-variate time serie, without a priori knowledge of the number of changepoints. Herein, we focus on detecting the most important changepoint (i.e. the one of largest magnitude), although we tested the method on a time-series that would have multiple changepoints (Fig S4). In order to test the performance, we detected the "real" changepoint on the whole time-series, and compare the changepoint found on the sub-sample with the "real" one. The distance to real changepoint served as performance diagnostic.

@@

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to real changepoint in commuper total number of samples analyzed, following (a) random sampling, (b) regular sampling, and (c) iterative sampling. Total number of samples was set between 5 and 30, out of the 74 initial time series.  \\label{fig:case_study_number_sample_change_biomass}'}
# Read data

@@

<!---- reexplain and compare the 3 methods (theoretical (Fig 3) vs case study (Fig 4), and how they relate (Fig 2)) ---->
Due to time or funding limitations, there is often a difference between the number of samples collected and the total number of samples that can later be processed. When the processing time is disproportionally higher than the collection time especially, a subsampling can be done prior to processing. A decision must then be taken of which subsamples to analyze. To address this question in the context of detecting changepoints, we tested three subsampling methods: subsampling at regular and random interval, and an iterative sampling approach (Fig. \ref{fig:conceptual}). We found that the iterative methods was systematically better at detecting changes than the two other methods, random subsampling being the least efficient (Figs. \ref{fig:simulation_approaches_plot}, \ref{fig:case_study_number_sample},S2,S3). Autocorrelation, variance, and shift size, had an effect on how many samples were needed to detect the shift, but did not changed this conclusion.  (Fig. \ref{fig:simulation_min_time_plot}).

@@

Our approach is applicable to a wide range of systems and questions, but it does have limitations. When less resources are needed for sample analysis, as opposed to collection, investigators will likely be able to process every sample. When sample analysis costs are high, then regular and interative sampling are better than random sampling (Figs. \ref{fig:simulation_approaches_plot},S2,S3). However, some questions require a continuous time series, or at least a regular sampling interval. For example, continuous subsampling of a time-series is required to detect critical slowing-down or early warning of shifts [@Frossard2015, @Doncaster2016]. However, critical slowing down does not necesseraly result in a shift, and a shift can occur without critical slowing down [@Spears2017]. Signs of critical slowing downs are important to understand and recognize because they provide potential early warnings [@Doncaster2016], but in terms of management, knowing most accurately the timing of a shift can have larger implication in adressing the driver. Thus, selecting a set number of samples or specific approach may also limit what future questions can be asked. However, more samples could be processed later to address these additional questions.

>>>>

Environmental monitoring is one of the core components to modern ecosystem research and management [@McDonald-Madden2010a, @White2019d, @Lindenmayer2020a]. Within an adaptive management framework, monitoring is needed for both learning about the system under study and assessing the effectiveness of management interventions [@Lovett2007]. Increasingly, long-term monitoring programs, like the Long Term Ecological Research (LTER) Network in the USA, are becoming available [@Maguran2010a]. However, environmental monitoring is still often expensive and time-consuming, especially when further processing is needed following sample collection (citations). Thus, for many fields there is a disparity between the amount of data that can be acquired and stored, and the ultimate number of samples that can be processed. Therefore, monitoring programs need to be designed in such a way to address the question of interest while using limited resources efficiently [@Legg2006a, @McDonald-Madden2010a, @Lindenmayer2020a].<!-- processing methods that could allow to extract the most information from that data. \rosalie{I feel like it could be a sensitive subject (we don't want to attack our peers and the way they do science), but there is definitely some truths. With some restoration work going on at the lab, there were some emails exchange about throwing away samples that were collected in 2015 and never looked at. If we could find the correct way to talk about this issue (without being judgemental, I'm sure it'd be interesting to many!).}

@@

Accounting for key considerations in the design of a monitoring program is necessary to detect long-term ecological change. The specifics of the monitoring program will determine the power with which a question of interest can be addressed. For example, @White2019d found that 72\% of vertebrate populations required at least 10 years of monitoring to detect significant changes in the population size over time. The specific number of years required depended on the species biology and the detection method used [@White2019d]. Other work has focused on the frequency of monitoring [@Wauchope2019a], showing that if a significant trend is observed from a dataset with limited temporal perspective, it is likely to reliably describe qualitatively the complete trend (increase or decline), but is unlikely to provide an accurate quantitative change in population. Further research investigated the impact of allocating monitoring resources spatially versus temporally [@Rhodes2011a, @Weiser2019a], and the ability to use citizen science to increase monitoring [@Weiser2020a]. Lastly, both the ecological and economical costs of failing to detect a true trend (type II error) have to be weighed against the risks of false (type I error) detection [@Mapstone1995a]. The approach to optimal sampling is thus a topical subject of research, as under limited budgets, monitoring programs need to be designed to be cost-effective [@Caughlan2001a, @Grantham2008b, @Bennett2016a].

@@

We detect changepoints with the function _e.divisive_ in the R package _ecp_ [@James2019]. There are several methods available to detect changepoint (reviewed in @James2014); _e.divisive_ is a divisive hierarchical estimation algorithm for multiple change point analysis. We chose this method because it is able to perform multiple change point analysis for both uni- and multi-variate time series, without a priori knowledge of the number of changepoints. Herein, we focus on detecting the most important changepoint (i.e. the one of largest magnitude), although we tested the method on a time-series that would have multiple changepoints (Fig S4). In order to test the performance, we detected the "real" changepoint on the whole time-series, and compare the changepoint found on the sub-sample with the "real" one. The distance to real changepoint served as performance diagnostic.

@@

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 3,fig.cap='Distance to true changepoint for total number of samples analyzed, following (a) random sampling, (b) regular sampling, and (c) iterative sampling. Total number of samples was set between 5 and 30, out of the 74 initial time series.  \\label{fig:case_study_number_sample_change_biomass}'}
# Read data

@@

<!---- reexplain and compare the 3 methods (theoretical (Fig 3) vs case study (Fig 4), and how they relate (Fig 2)) ---->
Due to time or funding limitations, there is often a difference between the number of samples collected and the total number of samples that can later be processed. When the processing time is disproportionately higher than the collection time especially, a subsampling can be done prior to processing. A decision must then be taken of which subsamples to analyze. To address this question in the context of detecting changepoints, we tested three subsampling methods: subsampling at regular and random interval, and an iterative sampling approach (Fig. \ref{fig:conceptual}). We found that the iterative methods was systematically better at detecting changes than the two other methods, random subsampling being the least efficient (Figs. \ref{fig:simulation_approaches_plot}, \ref{fig:case_study_number_sample},S2,S3). Autocorrelation, variance, and shift size, had an effect on how many samples were needed to detect the shift, but did not changed this conclusion.  (Fig. \ref{fig:simulation_min_time_plot}).

@@

Our approach is applicable to a wide range of systems and questions, but it does have limitations. When less resources are needed for sample analysis, as opposed to collection, investigators will likely be able to process every sample. When sample analysis costs are high, then regular and iterative sampling are better than random sampling (Figs. \ref{fig:simulation_approaches_plot},S2,S3). However, some questions require a continuous time series, or at least a regular sampling interval. For example, continuous subsampling of a time-series is required to detect critical slowing-down or early warning of shifts [@Frossard2015, @Doncaster2016]. However, critical slowing down does not necessarily result in a shift, and a shift can occur without critical slowing down [@Spears2017]. Signs of critical slowing downs are important to understand and recognize because they provide potential early warnings [@Doncaster2016], but in terms of management, knowing most accurately the timing of a shift can have larger implication in addressing the driver. Thus, selecting a set number of samples or specific approach may also limit what future questions can be asked. However, more samples could be processed later to address these additional questions.

### 2020-03-31 R/supp_material.Rmd
added figure caption and corrected small spelling mistakes

For each sampling approach, we wrote three custom functions:

@@

@@

source('R/sampling_random.R')
source('R/sampling_regular.R')
source('R/sampling_iterative.R')
source('R/build_time_series.R')
source('R/updated_param_sensitivity.R')
source('R/calc_min_time2.R')
require(tidyverse)

@@

@@

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 8}
library(ggplot2)
library(gridExtra)
p1 <- ggplot(myoutput[myoutput$target_cpt==1,], aes(final_n,diff_real, pch=factor(n2_init))) + 
  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
  geom_point(alpha = .3) +
  facet_wrap(~method_f) + theme_bw() +
  labs(title="a. First changepoint",x="", y="Distance to real changepoint") +
  theme(legend.position = c(0.93, 0.6),
        legend.background = element_rect(fill = "#ffffffaa", colour = NA),
        legend.title=element_text(size=7), 
        legend.text=element_text(size=8)) +
  guides(pch=guide_legend(title="Initial number\nof sample"))
p2 <- ggplot(myoutput[myoutput$target_cpt==2,], aes(final_n,diff_real, pch=factor(n2_init))) + 
  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
  geom_point(alpha = .3) +
  facet_wrap(~method_f) + theme_bw() +
  labs(title="b. Second changepoint",x="", y="Distance to real changepoint") +
  theme(legend.position = c(0.93, 0.6),
        legend.background = element_rect(fill = "#ffffffaa", colour = NA),
        legend.title=element_text(size=7), 
        legend.text=element_text(size=8)) +
  guides(pch=guide_legend(title="Initial number\nof sample"))
p3 <- ggplot(myoutput[myoutput$target_cpt==3,], aes(final_n,diff_real, pch=factor(n2_init))) + 
  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
  geom_point(alpha = .3) +
  facet_wrap(~method_f) + theme_bw() +
  labs(title="c. Third changepoint",x="", y="Distance to real changepoint") +
  theme(legend.position = c(0.93, 0.6),
        legend.background = element_rect(fill = "#ffffffaa", colour = NA),
        legend.title=element_text(size=7), 
        legend.text=element_text(size=8)) +
  guides(pch=guide_legend(title="Initial number\nof sample"))
grid.arrange(p1,p2,p3,nrow=3)
```

>>>>

We first detect the ``true" changepoint of a full length time series with the function _e.divisive_ in the R package _ecp_ [@James2019]. We focus on the changepoint with the largest magnitude, altough this package allows lower order changepoint detection as well (Fig. \ref{fig:case_study_other_changepoints}). We then subsample the full time series [@White2020] with different numbers of subsamples and different sampling approaches. For each sampling approach, we wrote a custom function:

@@

The subsampled time series are then compared to the full time series to assess the effectiveness of each subsampling approach.

@@

source('R/sampling_random.R')
source('R/build_time_series.R')
source('R/updated_param_sensitivity.R')
source('R/calc_min_time2.R')
require(tidyverse)

@@

p1 <- 
  ggplot(simulation_all %>% filter(Parameter == unique(simulation_all$Parameter)[1]), aes(Param, min_time)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept=0.05, lty=2) + 
  facet_wrap(~Parameter, scales="free_x", 
             labeller = label_glue('a. {Parameter}')) +
  ylim(10,52) +
  labs(x=c("sigma"), y= "Minimum number of samples required") + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        strip.background = element_rect(colour="transparent", fill="transparent"),
        strip.text.x = element_text(angle = 0, hjust = 0))
p2 <- ggplot(simulation_all %>% filter(Parameter == unique(simulation_all$Parameter)[2]), aes(Param, min_time)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 0.404, lty=2) +
  facet_wrap(~Parameter, scales="free_x", 
             labeller = label_glue('b. {Parameter}')) +
  ylim(10,52) +
  labs(x="phi", y= "") + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        strip.background = element_rect(colour="transparent", fill="transparent"),
        strip.text.x = element_text(angle = 0, hjust = 0))
p3 <- ggplot(simulation_all %>% filter(Parameter == unique(simulation_all$Parameter)[3]), aes(Param, min_time))  +
  geom_point() +
  geom_line() + 
  geom_vline(xintercept = 0.81, lty=2) +
  facet_wrap(~Parameter, scales="free_x", 
             labeller = label_glue('c. {Parameter}')) +
  ylim(10,52) +
  labs(x="shift size", y= "") + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        strip.background = element_rect(colour="transparent", fill="transparent"),
        strip.text.x = element_text(angle = 0, hjust = 0))

@@

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7.2, fig.height= 8,fig.cap='Distance to true (a) first changepoint (largest magnitude), (b) second changepoint, and (c) third changepoint for total number of samples analyzed, following random sampling, regular sampling, and iterative sampling. Total number of samples was set between 5 and 30, out of the 74 initial time series.  \\label{fig:case_study_other_changepoints}'}
library(ggplot2)
library(gridExtra)
p1 <- ggplot(myoutput[myoutput$target_cpt==1,], aes(final_n,diff_real, pch=factor(n2_init))) + 
  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
  geom_point(alpha = .3) +
  facet_wrap(~method_f) + theme_bw() +
  labs(title="a. First changepoint",x="", y="Distance to real changepoint") +
  theme(legend.position = c(0.93, 0.6),
        legend.background = element_rect(fill = "#ffffffaa", colour = NA),
        legend.title=element_text(size=7), 
        legend.text=element_text(size=8)) +
  guides(pch=guide_legend(title="Initial number\nof sample"))
p2 <- ggplot(myoutput[myoutput$target_cpt==2,], aes(final_n,diff_real, pch=factor(n2_init))) + 
  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
  geom_point(alpha = .3) +
  facet_wrap(~method_f) + theme_bw() +
  labs(title="b. Second changepoint",x="", y="Distance to real changepoint") +
  theme(legend.position = c(0.93, 0.6),
        legend.background = element_rect(fill = "#ffffffaa", colour = NA),
        legend.title=element_text(size=7), 
        legend.text=element_text(size=8)) +
  guides(pch=guide_legend(title="Initial number\nof sample"))
p3 <- ggplot(myoutput[myoutput$target_cpt==3,], aes(final_n,diff_real, pch=factor(n2_init))) + 
  geom_hline(aes(yintercept=0), col=adjustcolor("black", alpha.f = .4)) + 
  geom_point(alpha = .3) +
  facet_wrap(~method_f) + theme_bw() +
  labs(title="c. Third changepoint",x="", y="Distance to real changepoint") +
  theme(legend.position = c(0.93, 0.6),
        legend.background = element_rect(fill = "#ffffffaa", colour = NA),
        legend.title=element_text(size=7), 
        legend.text=element_text(size=8)) +
  guides(pch=guide_legend(title="Initial number\nof sample"))
grid.arrange(p1,p2,p3,nrow=3)
```
